Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2025-06-15 paper.qmd
grammar fixes and some larger overall comments

```{r}
#| label: fig-NLDR-variety
#| echo: false
#| fig-cap: "Eight different NLDR representations of the same data. Different techniques and different parameter choices are used. Researchers may have seen any of these in their analysis of this data, depending on their choice of method, or typical parameter choice. Would they make different decisions downstream in the analysis depending on which version seen? Which is the most accurate representation of the structure in high dimensions?"
#| fig-width: 8
#| fig-height: 4
#| out-width: 100%
#| fig-pos: H
# (a) UMAP (n_neighbors = 30, min_dist = 0.3), (b) UMAP (n_neighbors = 5, min_dist = 0.01), (c) UMAP (n_neighbors = 15, min_dist = 0.99), (f) tSNE (perplexity = 5), (g) tSNE (perplexity = 30), (l) TriMAP (n_inliers = 12, n_outliers = 4, n_random = 3), (q) PaCMAP (n_neighbors = 30, init = random, MN_ratio = 0.9, FP_ratio = 5)
nldr1 + nldr2 + nldr3 + nldr4 +
  nldr5 + nldr6 + nldr7 + nldr8 +
  plot_layout(ncol = 4)
```

@@

@@

<!-- - Connection between NLDR and MDS-->
Historically, \kD{} representations of \pD{} data have been computed using multidimensional scaling (MDS) [@kruskal1964], which includes principal components analysis (PCA) [@jolliffe2011] as a special case. (A contemporary comprehensive guide to MDS can be found in @borg2005.) The \kD{} representation can be considered to be a layout of points in \kD{}   produced by an embedding procedure that maps the data from \pD{}. In MDS, the \kD{} layout is constructed by minimizing a stress function that differences distances between points in \pD{} with potential distances between points in \kD{}. Various formulations of the stress function result in non-metric scaling [@saeed2018] and isomap [@silva2002]. Challenges in working with high-dimensional data, including visualization, are outlined in @johnstone2009.

Many new methods for NLDR have emerged in recent years, all designed to better capture specific structures potentially existing in \pD{}. Here we focus on five currently popular techniques, tSNE, UMAP, PHATE, TriMAP and PaCMAP. tNSE and UMAP can be considered to produce the \kD{}   minimizing the divergence between two distributions, where the distributions are modeling the inter-point distances. PHATE, TriMAP and PaCMAP are examples of diffusion processes [@coifman2005] spreading to capture geometric shapes, that include both global and local structure.

@@

We define the NLDR as a function $g\text{:}~ \mathbb{R}^{n\times p} \rightarrow \mathbb{R}^{n\times k}$, with (hyper-)~parameters $\mathbfit{\theta}$. The parameters, $\mathbfit{\theta}$, depend on the choice of $g$, and can be considered part of model fitting in the traditional sense. Common choices for $g$ include functions used in tSNE, UMAP, PHATE, TriMAP, PaCMAP, or MDS, although in theory any function that does this mapping is suitable.

@@

To ensure that the grid covers the range of data values a buffer parameter ($q$) is set as a proportion of the range. By default,  $q=0.1$. The buffer should be extending a full hexagon width ($a_1$) and height ($a_2$) beyond the data, in all directions. The lower left position where the grid starts is defined as $(s_1, s_2)$, and corresponds to the centroid of the lowest left hexagon, $C_{1}^{(2)} = (c_{11}, c_{12})$. This must be smaller than the minimum data value. Because it is one buffer unit, $q$ below the minimum data values, $s_1 = \text{-}q$ and $s_2 = \text{-}qr_2$.

@@

Observations are grouped into bins based on their nearest centroid. This produces a reduction in size of the data from $n$ to $m$, where $m\leq b$ (total number of bins). This can be defined using the function $u: \mathbb{R}^{n\times 2} \rightarrow \mathbb{R}^{m\times 2}$, where
$u(i) = \arg\min_{j = 1, \dots, b} \sqrt{(y_{i1} - C^{(2)}_{j1})^2 + (y_{i2} - C^{(2)}_{j2})^2}$, mapping observation $i$ into $H_h = \{i| u(i) = h\}$.

@@

Delaunay triangulation [@lee1980;@alb2024] is used to connect points so that edges indicate neighboring observations, in both the NLDR layout (@fig-NLDR-two-curvy (d)) and the \pD{} model representation. When the data has been binned the triangulation connectd centroids. The edges preserve the neighborhood information when the model is lifted into \pD{}.

When shapes are nonlinear in the NLDR layout, some edges could be long. It can also happen that distant centroids can be connected, particularly if clustering is present, which can result in long line segments. In order to generate a smooth surface in \gD{}, these long line segments should be removed when tuning the model fit.

@@

The last step is to lift the \gD{} model into \pD{} by computing \pD{} vectors that represent bin centroids. We use the \pD{} mean of the points in $H_h$ to map the centroid $C_{h}^{(2)} = (c_{h1}, c_{h2})$ to a point in \pD{}. Let the \pD{} mean be

$$C_{h}^{(p)} = \frac{1}{n_h}\sum_{i =1}^{n_h} x_i, h = {1, \dots, b; n_h > 0}.$$
Furthermore, line segments that exist in the \gD{} model generate line segments in \pD{} by connecting the \pD{} means of the corresponding \gD{} bin centroids. If additional long edges need to be removed, compute the edges in \pD{} and pruned any detected long edges to improve the accuracy. Once pruned, re-plot the \gD{} view to ensure it accurately captures the data.

@@

@@

where $n$ is the number of observations, $b$ is the number of bins, $n_h$ is the number of observations in $h^{th}$ bin, $p$ is the number of variables, $\mathbfit{x}_{hij}$ is the $j^{th}$ dimensional data of $i^{th}$ observation in $h^{th}$ hexagon. We can consider $e_{hj} = \sqrt{\sum_{j = 1}^{p} (\mathbfit{x}_{hij} - C^{(p)}_{hj})^2}$ to be the residual for each observation.

@@

```{r}
#| label: fig-p-d-error-in-2d-two-curvy
#| fig-cap: "The $4\\text{-}D$ model error in \\gD{} layout. Color indicates error ($e_{hj}$), dark color indicating high error and light indicates low error. Most large errors are distributed near the sparse end of the nonlinear cluster with dense corner."
#| out-width: 80%
#| fig-pos: H

@@

A new benefit of this fitted model is that it allows us to now predict a new observation's value in the NLDR, for any method. The steps are to determine the closest bin centroid in \pD{}, $C^{(p)}_{h}$ and predict it to be the centroid of this bin in \gD{}, $C^{(2)}_{h}$. This can be written as, let $z(i) = \arg\min_{j = 1, \dots, b} \sqrt{\sum_{v=1}^{p}(x_{iv} - C^{(p)}_{jv})^2}$, then the new observation $i$ falls in the hexagon, $H_h = \{i| z(i) = h\}$ and the corresponding \kD{} bin centroids, $C_{h}^{(2)} = (c_{h1}, c_{h2})$.

@@

The values $(s_1, \ s_2)$ define the position of the centroid of the bottom left hexagon. By default, this is at $s_1 = \text{-}q, s_2 = \text{-}qr_2$, where $q$ is the buffer bound the data. The choice of these values can have some effect on the distribution of bin counts. @fig-bins-two-curvy illustrates this. The distribution of bin counts for $s_1$ varying between $\text{-}0.1\,\text{--}\,0.0$ is shown. Generally, a more uniform distribution among these possibilities would indicate that the bins are reliably capturing the underlying distribution of observations.

@@

The default number of bins $b=b_1\times b_2$ is computed based on the sample size, by setting $b_1=n^{1/3}$, consistent with the Diaconis-Freedman rule [@freedman1981]. The value of $b_2$ is determined analytically by $b_1, q, r_2$. Values of $b_1$ between $2$ and $b_1 = \sqrt{n/r_2}$ are allowed. @fig-param-two-curvy (a) shows the effect of different choices of $b_1$ on the MSE of the fitted model.

#### Measurement of capturing the data shape in 2-D 
The area of a hexagon is defined as $A = 3\sqrt{3}l^2/2$ where $l$ is the side length of the hexagon. If we know $a_1$ and $a_2$, $l$ can be computed (see appendix). The density of a hexagon grid is calculated as $\sum^{h}_{i=1}n_h/A$ and the proportion is $\sum^{h}_{i=1}n_h/Ab$. The baseline proportion is the proportion density at the smallest possible value of $a_1$. The relative proportion density is the ratio of the observed proportion density to the baseline proportion density.

@@

By default, when assessing the choice of $b_1$, the total number of bins is measured by the number of **non-empty** bins. This more accurately reflects the hexagon grid relative the MSE than the full number of bins in the grid. It may also be beneficial to remove low count bins also, in the situation where data is clustered or stringy, where the observed data is sparse. In order to decide if this is necessary, you would examine the distribution of bin counts, or the density which puts the counts on a standard scale. If there is something of a gap at low values, this would suggest a potential value to use as a cutoff. Alternatively, one could choose to remove based on a percentile, the bins with density in the lowest 5% of all bins, for example. @fig-param-two-curvy (c) illustrates the effect on the model representation of removing bins below different percentages. Generally, we would urge caution in removing low count bins.

The benchmark value for removing low-density hexagons ranges between $0$ and $1$. When analyzing how these benchmark values influence model performance, it's essential to observe the change in MSE as the benchmark value increases (@fig-param-two-curvy (c)). The MSE shows a gradual decrease as the benchmark value goes from $1$ to $0$. Evaluating this rate of increase is important. If the increment is not considerable, the decision might lean towards retaining low-density hexagons.

@@

To understand how well the model fits the points whether it fits well, works better in some positions, or fails to match the overall pattern. It is important to link and brush the points with high model error in the \pD{} error plot, \gD{} layout, and the generated model overlay on the data in \pD{}.

@@

A particular pattern that we commonly see is that analysts tend to pick layouts with clusters that have big separations between them. When you examine their data in a tour, it is almost always that we see there are no big separations, and actually the suggested clusters are not even present. While we don't expect that analysts include animated gifs of tours in their papers, we should expect that any \gD{} representation adequately indicates the clustering that is present, and honestly show lack of separation or lack of clustering when it doesn't exist. It is important for analysts to have tools to select the accurate representation not the pretty but wrong representation.

To compare and assess a range of representations needs these items:

@@

The models are fitted to each these layouts. @fig-five-gau-projs a2, b2, c2 show the fitted models in a projection of the \fD{} space, taken from a tour. These clusters are fully \fD{} in nature, so we would expect the model to be a *crumpled sheet* that stretches in all four dimensions. This is what is mostly observed for tSNE and UMAP. The curious detail is that the model for PaCMAP is closer to a *pancake* in shape in every cluster! This single projection only shows this in three of the five clusters but if we examine a different projection the other clusters exhibit the pancake also. While we don't know what exactly causes this it is likely due to some ordering of points in the \gD{} PaCMAP layout that induces the flat model. One could imagine that if the method used principal components on all the data, that it might induce some ordering that would produce the flat model. If this were the reason, the pancaking would be the same in all clusters, but it is not: The pancake is visible in some clusters in some projections but in other clusters it is visible in different projections. It might be due to some ordering by nearest neighbors in a cluster. The PaCMAP documentation doesn't provide any helpful clues. That this happens, though, makes the PaCMAP layout inadequate for representing the high-dimensional data.

@@

@fig-one-dens_clust-error (a2, b2) shows \gD{} layout for tSNE created using the default (hyper-)~parameters. One would expect to see a rectangular shape if the curved sheet is flattened, but the layout is triangular. The other two displays show the residuals as a dot density plot (a1, b1), and a \gD{} projection of the data and the model from \fD{} (a3, b3). Using linked brushing between the plots, we can highlight points in the tSNE layout, and examine where they fall in the original \fD{}. The darker (maroon) points indicate points that have been highlighted by linking. In row a, the points at the top of the triangle are highlighted, and we can see these correspond to higher residuals, and also to all points at the low density end of the curved sheet. In row b, points at the lower left side of the triangle are highlighted which corresponds to smaller residuals and one corner of the sheet at the high density end of the curved sheet.

@@

In this data there are $2622$ single cells and $1000$ gene expressions (variables). Following their pre-processing, UMAP was performed on the first nine principal components. @fig-NLDR-variety shows this data using a variety of methods, and different (hyper-)~parameters. You can see that the result is wildly different depending on the choices. Layout a is a reproduction of the layout published in @chen2024. This layout suggests that the data has three very well separated clusters, each with an odd shape. The question is whether this accurately represents the cluster structure in the data, or whether they should have chosen b or c or d or e or f or g. This is what our new method can help with -- to decide which is the more accurate \gD{} representation of the cluster structure in the \pD{} data.

@@

The layout plot are colored to match the lines in the RMSE vs binwidth ($a_1$) plot. Layouts a, b and f fit the data better than c, d, e, and layout f appears to be the bets fit. @fig-clust-mnist shows this model in the data space in two projections from a tour. The data is curved in the $10\text{-}D$ space, and the fitted model captures this curve. The small clump of points in the \gD{} layout is highlighted in both displays. These are almost all inside the curve of the bulk of points and are sparsely located. The fact that they are packed together in the \gD{} layout is likely due to the handling of density differences by the NLDR.

The next step is to investigate the \gD{} to understand what information is learned from this representation. @fig-model-error-mnist summarizes this investigation. Plot a shows the layout with points colored by their residual value - darker color indicates larger residual and poor fit. The plots b, c, d, e show samples of hand-written digits taken from inside the colored boxes. Going from top to bottom around the curve shape we can see that the "1"s are drawn with from right slant to a left slant. The "1"s in d (black box) tend to have the extra up stroke but are quite varied in appearance. The "1"s shown in the plots labelled e correspond to points with big residuals. They can be seen to be more strangely drawn than the others. Overall, this \gD{} layout shows a useful way to summarize the variation in way "1"s are drawn.

@@

We have developed an approach to help assess and compare NLDR layouts, generated by different methods and (hyper-)~parameter choice(s). It depends on conceptualizing the \gD{} layout as a model, creating a wireframe representation of it that can be lifted into \pD{}. The fit is assessed by viewing the model in the data space, computing residuals and RMSE. Different layouts can be compared using the RMSE, and provides quantitative and objective methods for deciding on the most suitable NLDR layout to represent the \pD{} data. It also provides a way to predict the values of new \pD{} observations in the \gD{}, which could be useful for implementing uncertainty checks such as using training and testing samples.

>>>>

```{r}
#| label: fig-NLDR-variety
#| echo: false
#| fig-cap: "Eight different NLDR representations of the same data. Different methods and different parameter choices are used. Researchers may have seen any of these in their analysis of this data, depending on their choice of method, or typical parameter choice. Would they make different decisions downstream in the analysis depending on which version seen? Which is the most accurate representation of the structure in high dimensions?"
#| fig-width: 8
#| fig-height: 4
#| out-width: 100%
#| fig-pos: H
# (a) UMAP (n_neighbors = 30, min_dist = 0.3), (b) UMAP (n_neighbors = 5, min_dist = 0.01), (c) UMAP (n_neighbors = 15, min_dist = 0.99), (f) tSNE (perplexity = 5), (g) tSNE (perplexity = 30), (l) TriMAP (n_inliers = 12, n_outliers = 4, n_random = 3), (q) PaCMAP (n_neighbors = 30, init = random, MN_ratio = 0.9, FP_ratio = 5)
nldr1 + nldr2 + nldr3 + nldr4 +
  nldr5 + nldr6 + nldr7 + nldr8 +
  plot_layout(ncol = 4)
```

@@

<!-- Where are sections 4 & 5 in this description? -->

@@

<!-- - Connection between NLDR and MDS-->
Historically, low-dimensional (\kD{}) representations of high-dimensional (\pD{}) data have been computed using multidimensional scaling (MDS) [@kruskal1964], which includes principal components analysis (PCA) [@jolliffe2011] as a special case. (A contemporary comprehensive guide to MDS can be found in @borg2005.)
<!-- Does JASA allow footnotes? If so this is a footnote, otherwise the punctuation has gone a bit rogue - can't have an entire sentence in brackets surely --> The \kD{} representation can be considered to be a layout of points in \kD{}   produced by an embedding procedure that maps the data from \pD{}. In MDS, the \kD{} layout is constructed by minimizing a stress function that differences distances between points in \pD{} with potential distances between points in \kD{}. Various formulations of the stress function result in non-metric scaling [@saeed2018] and isomap [@silva2002]. Challenges in working with high-dimensional data, including visualization, are outlined in @johnstone2009.

<!-- Non-expert comment: This to me is a leap from MDS to NLDR. Is NLDR an MDS? I don't think so, but I'm missing the link -->
<!-- Suggested rewording: "To better capture specific structures potentially existing in \pD{}, many new methods for NLDR have emerged in recent years." -->
Many new methods for NLDR have emerged in recent years, all designed to better capture specific structures potentially existing in \pD{}. Here we focus on five currently popular techniques: tSNE, UMAP, PHATE, TriMAP and PaCMAP. Both tNSE and UMAP can be considered to produce the \kD{} representation by minimizing the divergence between two distributions, where the distributions are modeling the inter-point distances. PHATE, TriMAP and PaCMAP are examples of diffusion processes [@coifman2005] spreading to capture geometric shapes, that include both global and local structure.

@@

We define the NLDR as a function $g\text{:}~ \mathbb{R}^{n\times p} \rightarrow \mathbb{R}^{n\times k}$, with (hyper-)parameters $\mathbfit{\theta}$. These parameters, $\mathbfit{\theta}$, depend on the choice of $g$, and can be considered part of model fitting in the traditional sense. Common choices for $g$ include functions used in tSNE, UMAP, PHATE, TriMAP, PaCMAP, or MDS, although in theory any function that does this mapping is suitable.

@@

To ensure that the grid covers the range of data values a buffer parameter ($q$) is set as a proportion of the range. By default,  $q=0.1$. The buffer should be extending a full hexagon width ($a_1$) and height ($a_2$) beyond the data, in all directions. The lower left position where the grid starts is defined as $(s_1, s_2)$, and corresponds to the centroid of the lowest left hexagon, $C_{1}^{(2)} = (c_{11}, c_{12})$. This must be smaller than the minimum data value. Because it is one buffer unit, $q$ below the minimum data values, $s_1 = -q$ and $s_2 = -qr_2$.

@@

Observations are grouped into bins based on their nearest centroid. This produces a reduction in size of the data from $n$ to $m$, where $m\leq b$ (total number of bins). This can be defined using the function $u: \mathbb{R}^{n\times 2} \rightarrow \mathbb{R}^{m\times 2}$, where
$$u(i) = \arg\min_{j = 1, \dots, b} \sqrt{(y_{i1} - C^{(2)}_{j1})^2 + (y_{i2} - C^{(2)}_{j2})^2}$$, maps observation $i$ into $H_h = \{i| u(i) = h\}$.

<!-- I'm not thrilled about H_h but I guess it can stay because changing it everywhere is a pain -->

@@

Delaunay triangulation [@lee1980;@alb2024] is used to connect points so that edges indicate neighboring observations, in both the NLDR layout (@fig-NLDR-two-curvy (d)) and the \pD{} model representation. When the data has been binned the triangulation connects centroids. The edges preserve the neighborhood information from the \kD{} representation when the model is lifted into \pD{}.

When shapes are nonlinear in the NLDR layout, some edges could be long. It can also happen that distant centroids can be connected, particularly if clustering is present, which can result in long line segments. In order to generate a smooth surface in \gD{}, these long line segments should be removed when tuning the model fit. <!-- I don't think this is a thing anymore -->

@@

The last step is to lift the \gD{} model into \pD{} by computing \pD{} vectors that represent bin centroids. We use the \pD{} mean of the points in a given hexagon, $H_h$, denoted $C_{h}^{(p)}$, to map the centroid $C_{h}^{(2)} = (c_{h1}, c_{h2})$ to a point in \pD{}. Let the \pD{} mean be

$$C_{h}^{(p)} = \frac{1}{n_h}\sum_{i =1}^{n_h} x_i, h = {1, \dots, b; n_h > 0}.$$

Line segments that exist in the \gD{} model generate line segments in \pD{} by connecting the \pD{} means of the corresponding \gD{} bin centroids. If additional long edges need to be removed, compute the edges in \pD{} and prune any detected long edges to improve the accuracy. Once pruned, re-plot the \gD{} view to ensure it accurately captures the data. <!-- Do we still do this after the new binning system? -->

@@

<!-- In Figure 4, is the "sparse space" in the middle of each cluster (i.e. the holes) or is it the gap between clusters? If it's between clusters, I'd argue tSNE is making this gap smaller than in reality, which I don't think the caption currently conveys. -->

@@

where $n$ is the number of observations, $b$ is the number of bins, $n_h$ is the number of observations in $h^{th}$ bin, $p$ is the number of variables and $\mathbfit{x}_{hij}$ is the $j^{th}$ dimensional data of $i^{th}$ observation in $h^{th}$ hexagon. We can consider $e_{hj} = \sqrt{\sum_{j = 1}^{p} (\mathbfit{x}_{hij} - C^{(p)}_{hj})^2}$ to be the residual for each observation.

@@

<!-- Sparse end with dense corner to me is very hard to understand - sparse and dense are pretty much antonyms -->
```{r}
#| label: fig-p-d-error-in-2d-two-curvy
#| fig-cap: "The $4\\text{-}D$ model error in \\gD{} layout. Color indicates error ($e_{hj}$), dark color indicating high error and light indicates low error. Most large errors are distributed near the sparse end of the nonlinear cluster with dense corner."
#| out-width: 80%
#| fig-pos: H

@@

<!-- Does this really need an entire subsection? It's only 3 sentences long. -->

In this paradigm of model fitting, it is now possible to predict a new observation's value in the NLDR representation, independent of the choice of NLDR method. To do this, we first need to determine the closest bin centroid in \pD{}, $C^{(p)}_{h}$, to the new observation, which is,

$$z = \arg\min_{j = 1, \dots, b} \sqrt{\sum_{v=1}^{p}(x_{iv} - C^{(p)}_{jv})^2}.$$

Then, the predicted value is is the centroid of this bin in \gD{}, $C^{(2)}_{z} = (c_{z1}, c_{z2})$.

@@

The values $(s_1, \ s_2)$ define the position of the centroid of the bottom left hexagon. By default, this is at $s_1 = -q, s_2 = -qr_2$, where $q$ is the buffer bound the data. The choice of these values can have some effect on the distribution of bin counts which is seen in @fig-bins-two-curvy. <!-- Figure 6 doesn't seem to have any varying s_1 unless I'm missing something --> The distribution of bin counts for $s_1$ varying between $\text{-}0.1\,\text{--}\,0.0$ is shown. Generally, a more uniform distribution among these possibilities would indicate that the bins are reliably capturing the underlying distribution of observations.

@@

The default number of bins $b=b_1\times b_2$ is computed based on the sample size, by setting $b_1=n^{1/3}$, consistent with the Diaconis-Freedman rule [@freedman1981]. The value of $b_2$ is determined analytically by $b_1, q, r_2$ <!-- What is the formula for this? -->. Values of $b_1$ between $2$ and $b_1 = \sqrt{n/r_2}$ are allowed. @fig-param-two-curvy (a) shows the effect of different choices of $b_1$ on the MSE of the fitted model.

#### Measurement of capturing the data shape in 2-D

<!-- I'll be honest I've no idea what this section is -->
The area of a hexagon is defined as $A = 3\sqrt{3}l^2/2$ where $l$ is the side length of the hexagon. If we know $a_1$ and $a_2$, $l$ can be computed (see appendix). The density of a hexagon grid is calculated as $\sum^{h}_{i=1}n_h/A$ and the proportion is $\sum^{h}_{i=1}n_h/Ab$. The baseline proportion is the proportion density at the smallest possible value of $a_1$. The relative proportion density is the ratio of the observed proportion density to the baseline proportion density.

@@

By default, when assessing the choice of $b_1$, the total number of bins is measured by the number of **non-empty** bins. This more accurately reflects the hexagon grid relative the MSE than the full number of bins in the grid. It may be beneficial to remove low count bins also, such as when the data is clustered or stringy, or where the observed data is sparse. In order to decide if this is necessary, you would examine the distribution of bin counts, or the density which puts the counts on a standard scale. If there is something of a gap at low values, this would suggest a potential value to use as a cutoff. Alternatively, one could choose to remove based on a percentile, the bins with density in the lowest 5% of all bins, for example. @fig-param-two-curvy (c) illustrates the effect on the model representation of removing bins below different percentages. Generally, we would urge caution in removing low count bins.

The benchmark value for removing low-density hexagons ranges between $0$ and $1$, where $0$ corresponds to no bins being removed, and $1$ to 100% of bins being removed. When analyzing how these benchmark values influence model performance, it's essential to observe the change in MSE as the benchmark value increases (@fig-param-two-curvy (c)). The MSE shows a gradual decrease as the benchmark value goes from $1$ to $0$. Evaluating this rate of increase is important. If the increment is not considerable, the decision might lean towards retaining low-density hexagons.

@@

To understand how well the model fits the points whether it fits well, works better in some positions, or fails to match the overall pattern, it is important to link and brush the points with high model error in the \pD{} error plot, \gD{} layout, and the generated model overlay on the data in \pD{}.

@@

A particular pattern that we commonly see is that analysts tend to pick layouts with clusters that have big separations between them. When you examine their data in a tour, it is almost always that we see there are no big separations, and actually often the suggested clusters are not even present. While we don't expect that analysts include animated gifs of tours in their papers, we should expect that any \gD{} representation adequately indicates the clustering that is present, and honestly show lack of separation or lack of clustering when it doesn't exist. It is important for analysts to have tools to select the accurate representation not the pretty but wrong representation.

To compare and assess a range of representations an analyst needs:

@@

The models are fitted to each these layouts. @fig-five-gau-projs a2, b2, c2 show the fitted models in a projection of the \fD{} space, taken from a tour. These clusters are fully \fD{} in nature, so we would expect the model to be a *crumpled sheet* that stretches in all four dimensions. This is what is mostly observed for tSNE and UMAP. The curious detail is that the model for PaCMAP is closer to a *pancake* in shape in every cluster! This single projection only shows this in three of the five clusters but if we examine a different projection the other clusters exhibit the pancake also. While we don't know what exactly causes this, it is likely due to some ordering of points in the \gD{} PaCMAP layout that induces the flat model. One could imagine that if the method used principal components on all the data, that it might induce some ordering that would produce the flat model. If this were the reason, the pancaking would be the same in all clusters, but it is not: The pancake is visible in some clusters in some projections but in other clusters it is visible in different projections. It might be due to some ordering by nearest neighbors in a cluster. The PaCMAP documentation doesn't provide any helpful clues. That this happens, though, makes the PaCMAP layout inadequate for representing the high-dimensional data.

@@

@fig-one-dens_clust-error (a2, b2) shows a \gD{} layout for tSNE created using the default (hyper-)~parameters. One would expect to see a rectangular shape if the curved sheet is flattened, but the layout is triangular. The other two displays show the residuals as a dot density plot (a1, b1), and a \gD{} projection of the data and the model from \fD{} (a3, b3). Using linked brushing between the plots, we can highlight points in the tSNE layout, and examine where they fall in the original \fD{}. The darker (maroon) points indicate points that have been highlighted by linking. In row a, the points at the top of the triangle are highlighted, and we can see these correspond to higher residuals, and also to all points at the low density end of the curved sheet. In row b, points at the lower left side of the triangle are highlighted which corresponds to smaller residuals and one corner of the sheet at the high density end of the curved sheet.

@@

In this data there are $2622$ single cells and $1000$ gene expressions (variables). Following their pre-processing, UMAP was performed on the first nine principal components. @fig-NLDR-variety shows this data using a variety of methods, and different (hyper-)~parameters. <!-- I think some of these panels aren't umap - if so saying "umap was performed on the first nine principal components" is not necessarily correct. I think its what the publication did but not what is in our figure --> You can see that the result is wildly different depending on the choices. Layout a is a reproduction of the layout published in @chen2024. This layout suggests that the data has three very well separated clusters, each with an odd shape. The question is whether this accurately represents the cluster structure in the data, or whether they should have chosen b or c or d or e or f or g. This is what our new method can help with -- to decide which is the more accurate \gD{} representation of the cluster structure in the \pD{} data.

@@

The layout plots are colored to match the lines in the RMSE vs binwidth ($a_1$) plot. Layouts a, b and f fit the data better than c, d, e, and layout f appears to be the bets fit. @fig-clust-mnist shows this model in the data space in two projections from a tour. The data is curved in the $10\text{-}D$ space, and the fitted model captures this curve. The small clump of points in the \gD{} layout is highlighted in both displays. These are almost all inside the curve of the bulk of points and are sparsely located. The fact that they are packed together in the \gD{} layout is likely due to the handling of density differences by the NLDR.

The next step is to investigate the \gD{} layout to understand what information is learned from this representation. @fig-model-error-mnist summarizes this investigation. <!-- I understand the convenience of b-f all lined up but I did go looking for a red box. With enough reading I know why it's not there but I wonder if this can be improved --> Plot a shows the layout with points colored by their residual value - darker color indicates larger residual and poor fit. The plots b, c, d, e show samples of hand-written digits taken from inside the colored boxes. Going from top to bottom around the curve shape we can see that the "1"s are drawn with from right slant to a left slant. The "1"s in d (black box) tend to have the extra up stroke but are quite varied in appearance. The "1"s shown in the plots labelled e correspond to points with big residuals. They can be seen to be more strangely drawn than the others. Overall, this \gD{} layout shows a useful way to summarize the variation in way "1"s are drawn.

@@

We have developed an approach to help assess and compare NLDR layouts, generated by different methods and (hyper-)~parameter choice(s). It depends on conceptualizing the \gD{} layout as a model, allowing for the creation of a wireframe representation of the model that can be lifted into \pD{}. The fit is assessed by viewing the model in the data space, computing residuals and RMSE. Different layouts can be compared using the RMSE, and provides quantitative and objective methods for deciding on the most suitable NLDR layout to represent the \pD{} data. It also provides a way to predict the values of new \pD{} observations in the \gD{}, which could be useful for implementing uncertainty checks such as using training and testing samples.
